---
title: "Reproducible Research: Peer Assessment 1"
author: "kalyan"
date: "May 17, 2015"
output: html_document
---
        
#Loading and Processsing
        
```{r, echo=TRUE}
activity <- read.csv("activity.csv", colClasses = c("numeric", "character", 
                                                    "numeric"))
head(activity)
```

```{r, echo=TRUE}
names(activity)
```

```{r, echo=TRUE}
library(lattice)
```

```{r, echo=TRUE}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

#Mean total number of steps taken per day?

Two approaches can be followed for this

First one is using the aggregate function

```{r, echo=TRUE}
StepsTotal <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)
```

Plot histogram of the total number of steps taken each day:
        
```{r, echo=TRUE}
hist(StepsTotal$steps, main = "Total steps by day", xlab = "day", col = "orange")
```

Mean and median of steps:
        
```{r, echo=TRUE}
mean(StepsTotal$steps)
```

```{r, echo=TRUE}
median(StepsTotal$steps)
```

The second approach is to make a data frame first with the values I needed. I create a data frame with the days and the total number of steps per day

```{r, echo=TRUE}
steps <- rep(NA, 61)
day <- rep("NA", 61)
stepsday <- tapply(activity$steps, activity$date, sum, na.rm = T)
length(stepsday)
```

```{r, echo=TRUE}
for (i in 1:61) {
    steps[i] <- stepsday[[i]]
    day[i] <- names(stepsday)[i]
}
```

data frame:

```{r, echo=TRUE}
df <- data.frame(day, steps)
head(df)
```

Time for the histogram..

```{r, echo=TRUE}
hist(df$steps, main = "Total steps by day", xlab = "day", col = "pink")
```

This second histogram is a bit differnt because there are days with no data for the steps.

#What is the average daily activity pattern?
Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
time_series <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
```

Generate the plot

```{r, echo=TRUE}
plot(row.names(time_series), time_series, type = "l", xlab = "5-min interval", 
    ylab = "Average across all Days", main = "Average number of steps taken", 
    col = "red")
```      

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 

```{r, echo=TRUE}
max_interval <- which.max(time_series)
names(max_interval)
```

#Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo=TRUE}
activity_NA <- sum(is.na(activity))
activity_NA
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval,etc

Fist Na replaced by mean in 5 min interval

```{r, echo=TRUE}
StepsAverage <- aggregate(steps ~ interval, data = activity, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(activity)) {
    obs <- activity[i, ]
    if (is.na(obs$steps)) {
        steps <- subset(StepsAverage, interval == obs$interval)$steps
    } else {
        steps <- obs$steps
    }
    fillNA <- c(fillNA, steps)
}
```    

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo=TRUE}
new_activity <- activity
new_activity$steps <- fillNA
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r, echo=TRUE}
StepsTotal2 <- aggregate(steps ~ date, data = new_activity, sum, na.rm = TRUE)
```

The Histogram is

```{r, echo=TRUE}
hist(StepsTotal2$steps, main = "Total steps by day", xlab = "day", col = "red")
```

The mean and median are ...

```{r, echo=TRUE}
mean(StepsTotal2$steps)
```

```{r, echo=TRUE}
median(StepsTotal2$steps)
```

After replacing, mean remains the same, but the median varies

#Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels  weekday and weekend indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
day <- weekdays(activity$date)
daylevel <- vector()
for (i in 1:nrow(activity)) {
    if (day[i] == "Saturday") {
        daylevel[i] <- "Weekend"
    } else if (day[i] == "Sunday") {
        daylevel[i] <- "Weekend"
    } else {
        daylevel[i] <- "Weekday"
    }
}
activity$daylevel <- daylevel
activity$daylevel <- factor(activity$daylevel)

stepsByDay <- aggregate(steps ~ interval + daylevel, data = activity, mean)
names(stepsByDay) <- c("interval", "daylevel", "steps")
```

Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

```{r, echo=TRUE}
xyplot(steps ~ interval | daylevel, stepsByDay, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")
```